---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# arcgis

<!-- badges: start -->
<!-- badges: end -->

The goal of `{arcgis}` is to provide an R interface to the ArcGIS REST API 

## Installation

You can install the development version of arcgis like so:


``` r
remotes::install_github("R-ArcGIS/api-interface")
```

## Feature Layers

Read in data from a feature layer. arcgis implements a `FeatureLayer` class object which is used to control how we read data from a Feature Layer service. 

Create a `FeatureLayer` with `feature_layer()` which takes the url of the feature layer as its only argument. 

```{r example}
library(arcgis)

# define the feature layer url
furl <- "https://services.arcgis.com/P3ePLMYs2RVChkJx/ArcGIS/rest/services/USA_Counties_Generalized_Boundaries/FeatureServer/0"

# create a feature layer
county_fl <- feature_layer(furl)

county_fl
```

`FeatureLayer` objects are lists that contain metadata from the FeatureServer. The data that is used to create the `FeatureLayer` is taken directly from the url when displayed in json format. `FeatureLayer`s are intended to work somewhat similarly to a lazy table where data is only brought into memory when explicitly requested. 

The `collect_layer()` function will fetch all data for a layer from the REST endpoint and write it to the appropriate type—either an sf object or a data frame. 

```{r cars}
county_sf <- collect_layer(county_fl)

county_sf
```

We can check to see what fields and types are present in our feature layer using `list_fields()`. 

```{r}
list_fields(county_fl)
```

You can provide a more detailed query to the layer using `query_layer()`. 


```{r}
res <- query_layer(
  county_fl, 
  where = "STATE_NAME = 'California' and POPULATION >= 30000",
  fields = c("FIPS", "STATE_NAME", "SQMI", "POPULATION")
)

res
```

Additionally, you can provide sf objects as spatial filters based on a predicate function.  Possible options are "intersects", "contains", "crosses", "overlaps", "touches", and "within".

In this example we create a polygon representing the bounding box of the West Coast of the US to act as our spatial filter. 

```{r}
west_coast_bbox <- sf::st_as_sfc(
  sf::st_bbox(c(xmin = -125, ymin = 32, xmax = -114, ymax = 49))
)

west_coast_bbox
```

We can then provide it as input to the `filter_geom` argument to act as a spatial filter. 

```{r}
res <- query_layer(
  county_fl, 
  filter_geom = west_coast_bbox
)

plot(res$geometry)
plot(west_coast_bbox, lty = 3, add = TRUE)
```

For manual control of the parameters that will be sent to the REST API you can use `update_params()`. This function allows you to built up your query iteratively and programmatically without executing the query right away. Use it in conjunction with `collect_layer()` for a lazy-ish process.

```{r}
fl_query <- update_params(
  county_fl, 
  where = "STATE_NAME = 'New Hampshire'",
  outFields = "STATE_NAME,POPULATION"
)

fl_query
```

Notice the appended query. You can execute the query and bring the results into memory using `collect_layer()`.

```{r}
collect_layer(fl_query)
```

## Interfacing with `{arcgisbinding}`


- to do....


### dplyr-like syntax

You may not need all of these fields or features in your R session. If that's the case, it is more effective to bring in only a subset of the data that you need. We can limit what is brought into memory by using `select()` and `filter()` functions to build up a query that can be sent the rest API. 

The `select()` `FeatureLayer` method supports using `{tidyselect}` functions except the `where()` function.

```{r}
county_fl |> 
  select(contains("STATE"))
```

Now that we've ran `select()` on the `FeatureLayer` query parameters are printed out. The parameters will be sent along when the results are `collect()`ed. 

You can further subset your data by using `filter()`. Note that filtering is fairly limited ([see API documentation for more](https://developers.arcgis.com/rest/services-reference/enterprise/query-feature-service-layer-.htm#GUID-BC2AD141-3386-49FB-AA09-FF341145F614)).

```{r}
county_query <- county_fl |> 
  select(starts_with("STATE")) |> 
  filter(STATE_NAME == "Alabama")

county_query
```

Bring it into memory with `collect()`.

```{r}
res <- collect(county_query)
res
```


### Other `FeatureLayer` functions

You can add any parameter that you'd like that is supported by the query parameters [listed in the documentation](https://developers.arcgis.com/rest/services-reference/enterprise/query-feature-service-layer-.htm#GUID-BC2AD141-3386-49FB-AA09-FF341145F614) with the using the `update_params()` function.


```{r}
county_query |> 
  update_params(returnIdsOnly = "true") |> 
  collect()
```

Maybe, you want to clear our your query parameters, use `clear_query()`.

```{r}
clear_query(county_query)
```

To preview only a few features from the feature layer use `head()`

```{r}
head(county_query)
```


## `FeatureServers`


FeatureServers contain one or more `FeatureLayers` or `Tables`. 

```{r}
ft_srv <- feature_server(
  "https://services2.arcgis.com/j80Jz20at6Bi0thr/ArcGIS/rest/services/List_of_Providers/FeatureServer/"
  )

ft_srv
```

Access elements by ID with `get_layer()`

```{r}
get_layer(ft_srv, 27)
```

Get a list of all elements.

```{r}
get_all_layers(ft_srv)
```

## Conversion to Esri JSON


sf objects can be converted to Esri JSON representation. These can be Geometry Objects, FeatureSets, or arrays of Features. 

`st_as_geometry()` works with objects of class `sfg` and creates a Geometry Object. `st_as_featureset()` creates a FeatureSet from an sf or sfc object. sfc objects will have 0 length attributes. `st_as_features()` works with an sf or sfc object. 


```{r}
st_as_geometry(res[["geometry"]][[30]]) |> 
  # print nicely with jsonify
  jsonify::pretty_json()
```

```{r}
st_as_featureset(res[30,]) |> 
  jsonify::pretty_json()
```
